---
title: "A2-ex1&4"
author: "Paola Vazquez Castillo"
date: "10/12/2020"
output: pdf_document
---

```{r include=FALSE}
library(tidyverse)
library(DescTools)
library(wpp2019)
```

```{r include=FALSE}
setwd("~/Mirror/EDSD/Basic Math/Assignments/2")
```

### Exercise 1

A simplified model of population growth can be expressed as:
$$P(t)=P(t_o)e^{r(t-t_o)} \tag{1}$$

where $P(t)$ is the population size at time $t$, $P(t_0)$ is the population at initial time $t_0$ and $r$ is the constant growth rate. If interested, see Keyfitz and Caswel (2005, Chap. 1) for a detailed discussion.

a. Suppose the growth rate is $r = 0.01$. In how many years the population will double its size with respect to $P(t_0)$?

[Sol.]

From (1),

$$\frac{P(t)}{P(t_0)}=e^{r(t-t_0)} \Rightarrow ln(\frac{P(t)}{P(t_0)})=r(t-t_0) \Rightarrow \frac{ln(\frac{P(t)}{P(t_0)})}{r}=t-t_0 \tag{a}$$

where, $t-t_0$ is the time elapsed. On the other hand, having the double of the initial population means: 

$$\frac{P(t)}{P(t_0)}=\frac{2P(t_0)}{P(t_0)}=2 \tag{b}$$

Therefore, substituting $(**)$ in $(*)$:
$$t-t_0=\frac{ln(2)}{r} \tag{c}$$
Note that $eq.(c)$ is the doubling time for any r. Substituting the data of this problem.

$$\Rightarrow t-t_0=\frac{ln(2)}{0.01}=69.31$$

b. Calculate the growth rate if the population doubles its size in 60 years; if the population triples its size in 100 years; and if the population reduces to 50% of its size in 50 years.

[Sol.]

From (1),

$$r=\frac{ln(\frac{P(t)}{P(t_0)})}{t-t_o} \tag{d}$$
Therefore, substituting in (d):

b1. $r=\frac{ln(2)}{60}\approx 0.0115$

b2. $r=\frac{ln(3)}{100}\approx 0.0110$

b3. $r=\frac{ln(\frac{1}{2})}{50}\approx -0.0139$


c. Select three countries of your choice from three different continents. Using data from the World Population Prospects (United Nations, 2017) and applying (1), estimate the growth rate of each of their populations between 1950 and 2015 (ignoring intermediate years).

Selection of random* countries for study.

*Based on two lists one of the countries where the members of the team are native and the other from the Asian Tigers. Italy is the European country chosen.

```{r echo=FALSE}
# One American Country 
list<-c("Mexico","Canada","Colombia","Brasil")
set.seed(333)
sample(list,1)

# One Asian Country (one among the Asian Tigers)
list2<-c("Hong Kong","Singapore","South Korea", "Taiwan")
set.seed(222)
sample(list2,1)
```


```{r include= FALSE}
# Read data
data("pop")

# Data of interest for ex c and d
WD<-pop%>%
  filter(country_code==344 | country_code==380| country_code==124)%>%
  select(name,`1950`,`1990`,`2015`) 
```

[Sol.]
```{r }
# c rates 1950-2015
# HongKong
rM1950<- (log(WD[1,4]/WD[1,2]))/(2015-1950)
rM1950
# Italy
rI1950<- (log(WD[2,4]/WD[2,2]))/(2015-1950)
rI1950
#Canada
rC1950<- (log(WD[3,4]/WD[3,2]))/(2015-1950)
rC1950
```
d. Repeat the same procedure, but in this case compute the population growth between 1990 and 2015.

[Sol.]
```{r}
#d rates 1990-2015
# HongKong
rM1990<- (log(WD[1,4]/WD[1,3]))/(2015-1990)
rM1990
# Italy
rI1990<- (log(WD[2,4]/WD[2,3]))/(2015-1990)
rI1990
# Canada
rC1990<- (log(WD[3,4]/WD[3,3]))/(2015-1990)
rC1990
```

e. For each of the three countries, do a single plot with the observed population counts from 1950 until 2015, adding two lines with each of the estimated populations using the results from the two previous questions (one for 1950–2015, and another for 1990–2015).

```{r echo=FALSE}
#e plot 
Mex<-pop%>%
  filter(country_code==344) %>% 
  gather(year,pop,3:17)%>%
  select(year,pop)

Ita<-pop%>%
  filter(country_code==380) %>% 
  gather(year,pop,3:17)%>%
  select(year,pop)

Can<-pop%>%
  filter(country_code==124) %>% 
  gather(year,pop,3:17)%>%
  select(year,pop)
```
```{r echo=FALSE}
# Years of estimation
yrs<-c(seq(1950,2015,5))

# Setting initial populations 
PM1950<-Mex[1,2]
PI1950<-Ita[1,2]
PC1950<-Can[1,2]

PM1990<-Mex[9,2]
PI1990<-Ita[9,2]
PC1990<-Can[9,2]

Mexr1950<-c(PM1950*(exp(rM1950*(yrs-1950))))
Mexr1990<-c(PM1950*(exp(rM1990*(yrs-1950))))

Itar1950<-c(PI1950*(exp(rI1950*(yrs-1950))))
Itar1990<-c(PI1950*(exp(rI1990*(yrs-1950))))

Canr1950<-c(PC1950*(exp(rC1950*(yrs-1950))))
Canr1990<-c(PC1950*(exp(rC1990*(yrs-1950))))

```

[Sol.]


```{r echo=FALSE}
# Hong Kong
matplot(c(seq(1950,2015,5)),cbind(Mex$pop[1:14],
              Mexr1950,
              Mexr1990),
        type="l",
        ylab = "Population (in thousands)",
        xlab = "Year",
        lty = c(1,2,2),
        col =c("black","red","blue"),
        main="Population estimates* in Hong Kong,1950-2015")
legend("topleft",
       legend = c("Observed","1950-2015 growth rate","1990-2015 growth rate"),
       col= c("black","red","blue"),
       lty= c(1,2,2),
       box.lty=0,
       cex = 0.8)
mtext("*Estimates for 2 different growth rates",side=1,line=3,outer=F,
      cex=0.8,adj=1,at=c(1968),font=3)
mtext("Source: Own calculations from World Population Prospects, 2019 revision",side=1,line=4,outer=F,
      cex=0.8,adj=1,at=c(1997),font=3)

```

```{r echo=FALSE}
# Italy
matplot(c(seq(1950,2015,5)),cbind(Ita$pop[1:14],
                                  Itar1950,
                                  Itar1990),
        type="l",
        ylab = "Population (in thousands)",
        xlab = "Year",
        lty = c(1,2,2),
        col =c("black","red","blue"),
        main="Population estimates* in Italy,1950-2015")
legend("topleft",
       legend = c("Observed","1950-2015 growth rate","1990-2015 growth rate"),
       col= c("black","red","blue"),
       lty= c(1,2,2),
       box.lty=0,
       cex = 0.8)
mtext("*Estimates for 2 different growth rates",side=1,line=3,outer=F,
      cex=0.8,adj=1,at=c(1965.5),font=3)
mtext("Source: Own calculations from World Population Prospects, 2019 revision",side=1,line=4,outer=F,
      cex=0.8,adj=1,at=c(1991),font=3)


```

```{r echo=FALSE}
# Canada
matplot(c(seq(1950,2015,5)),cbind(Can$pop[1:14],
                                  Canr1950,
                                  Canr1990),
        type="l",
        ylab = "Population (in thousands)",
        xlab = "Year",
        lty = c(1,2,2),
        col =c("black","red","blue"),
        main="Population estimates* in Canada,1950-2015")
legend("topleft",
       legend = c("Observed","1950-2015 growth rate","1990-2015 growth rate"),
       col= c("black","red","blue"),
       lty= c(1,2,2),
       box.lty=0,
       cex = 0.8)
mtext("*Estimates for 2 different growth rates",side=1,line=3,outer=F,
      cex=0.8,adj=1,at=c(1965.5),font=3)
mtext("Source: Own calculations from World Population Prospects, 2019 revision",side=1,line=4,outer=F,
      cex=0.8,adj=1,at=c(1991),font=3)

```


f. Briefly discuss the differences among the three countries, and which of the estimated growth rates are better according to the graphs.

[Sol.]

From the graph we can note that in the three countries, the blue line, which is considering the growth rates form 1990 to 2015, produce lower population estimates compared to the one considering the 1950-2015 rate. This makes sense considering that these three countries experimented the second phase of the demographic somewhen between 1950-1990, meaning that the fertility rates lowered within the period producing lower population estimates. As there was a disrupting event, it is better considering the rates calculated with data form the begining and the end of the estimation period.

Comparing the three countries, they differ the most on their population sizes, but also in the speed of their growth (measured with the rate of growth), which is also a phenomena that can be appreciated in the graph by comparing the slopes of the lines. 


### Exercise 4

A man aged 30 marries a woman aged 25. Express in symbols the expected number of years they will both be alive. If $\mu_x$ is constant and equal to $0.02$, evaluate the symbolic expression.

[Sol.]

The time the man will be alive is $e_{30}^M$ and the woman $e_{25}^F$.We also know that $e_x=\int_x^\infty e^{-\int_x^a\mu(y)dy}da$. Therefore,

$$e_{30}^M=\int_{30}^{\infty}e^{-\int_x^a \mu(y)dy}da=\int_{30}^{\infty}e^{-\int_{30}^a (0.02)dy}da=\int_{30}^\infty e^{-(0.02)y|_{30}^a}da=\int_{30}^\infty e^{-[(0.02)a-(0.02)*30]}da=$$

$$\int_{30}^\infty e^{-0.02a}*e^{0.6}da=e^{0.6}\int_{30}^\infty e^{-0.02a}da=e^{0.6}\frac{-1}{(0.02)}e^{-0.02a|_{30}^\infty}=0-e^{0.6}\frac{-1}{(0.02)}e^{-0.02*30}=\frac{1}{0.02}=50$$
Conversely,

$$e_{25}^F=\int_{25}^{\infty}e^{-\int_{25}^a \mu(y)dy}da=\int_{25}^{\infty}e^{-\int_{25}^a (0.02)dy}da=\int_{25}^\infty e^{-(0.02)y|_{25}^a}da=\int_{25}^\infty e^{-[(0.02)a-(0.02)*25]}da=$$
$$\int_{25}^\infty e^{-0.02a}*e^{0.5}da=e^{0.5}\int_{25}^\infty e^{-0.02a}da=e^{0.5}\frac{-1}{(0.02)}e^{-0.02a|_{25}^\infty}=0-e^{0.5}\frac{-1}{(0.02)}e^{-0.02*25}=\frac{1}{0.02}=50$$

Therefore,

the expected number of years they will both be alive is in average 50 years each one.This is an empiric example that, when the mortality force remains constant the life expectancy is the same at every age, because there are not differences in the "way" people from age to age dies (or among sexes). This is an unrealistic scenario that can only happen if there is an homogenous population without biological aging or mortality risk sex/age/social differences.


